Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2020-09-06 Paper/manuscript_v2/manuscript_07_02.tex
Unify 'lengthscale' word

The accuracy and speed of the HSGP model depends on several interrelated factors, most notably on the number of basis functions and on the boundary condition of the Laplace eigenfunctions. Furthermore, appropriate values for these factors will depend on the degree of non-linearity of the function to be estimated, which is in turn is characterized by the lengthscale of the covariance function. In this section, we analyze the effects of the number of basis functions and the boundary condition on the approximation accuracy. We present recommendations on how they should be chosen and diagnostics to check the accuracy of the obtained approximation.

Ultimately, these recommendations are based on the relationships among the number of basis functions, the boundary condition and the lengthscale of the function, which depend on the particular choice of the kernel function. In this work we investigate these relationships for the square exponential and the Mat{\'e}rn ($\nu$=3/2) covariance functions in the present section, and for the periodic squared exponential covariance function in Appendix~\ref{ch5_sec_periodic}. For other kernels, the relationships will be slightly different depending on the smoothness or wigglyness of the covariance function.

@@

This inequality can be satisfied for arbritrary small $\epsilon$ provided that $L$ and $m$ are sufficiently large \cite[Theorem 1 and 4]{solin2018hilbert}. 
The specific number of basis functions $m$ needed depends on the degree of non-linearity of the function to be estimated, that is on its lengthscale $\ell$, which constitutes a hyperparameter of the G. The approximation also depends on the boundary $L$ (see equations (\ref{eq_eigenvalue}), (\ref{eq_eigenfunction}), (\ref{eq_eigenvalue_multi}) and (\ref{eq_eigenfunction_multi})), which will affect its accuracy especially near the boundaries. As we will see later on, $L$ will also influence the number of basis functions required in the approximation.

@@

We start by illustrating how the number of basis functions $m$ and boundary factor $c$ influence the accuracy of the HSGP approximations individually. For this purpose, a set of noisy observations are drawn from an exact GP model with a squared exponential covariance function of lengthscale $\ell=0.3$ and marginal variance $\alpha=1$, using input values from the zero-mean input domain with half-range $S=1$. Several HSGP models with varying $m$ and $c$ are fitted to this data. In this example, the lengthscale and marginal variance parameters used in the HSGPs are fixed to the true values of the data-generating model. 
Figures~\ref{fig1_Post_J} and \ref{fig2_Post_L} illustrate the individual effects of $m$ and $c$, respectively, on the posterior predictive mean and standard deviation of the estimated function as well as on the covariance function itself. For $c$ fixed to a sufficiently large value, Figure \ref{fig1_Post_J} shows clearly how $m$ affects the accuracy on the approximation for both the posterior mean or uncertainty. It is seen that if the number of basis functions $m$ is too small, the estimated function tend to be overly smooth because the necessary high frequency components are missing. In general, the higher the degree of wigglyness of the function to be estimated, the larger number of  basis functions will be required. If $m$ is fixed to a sufficiently large value, Figure~\ref{fig2_Post_L} shows that $c$ affects the approximation mainly near the boundaries in mean, but along the whole domain in the standard deviation. The approximation error for the variance tends to be bigger for the variance than for the mean.

@@

Next, we analyze how the interaction effects between $m$ and $c$ affects the quality of the approximation. The lengthscale and marginal variance of the covariance function will no longer be fixed but instead we compute the joint posterior distribution of the function values and the hyperparameters using the dynamic HMC algorithm implemented in Stan \citep{carpenter2017stan,betancourt2017conceptual} for both the exact GP and the HSGP models. Figure~\ref{fig3_Post_part1} shows the posterior predictive mean and standard deviation of the function as well as the covariance function obtained after fitting the model for varying $m$ and $c$. Figure \ref{fig4_MSE_vs_J} shows the root mean square error (RMSE) of the HSGP models computed against the exact GP model. Figure \ref{fig5_lscale_vs_J} shows the estimated lengthscale and marginal variance for the exact GP model and the HSGP models. Looking at the RMSEs in Figure \ref{fig4_MSE_vs_J}, we can conclude that the optimal choice in terms of precision and computation time for this example would be $m = 15$ basis functions and a boundary factor between $c = 1.5$ and $c = 2.5$. Further, the less conservative choice of $m = 10$ and $c = 1.5$  could also produce a sufficiently accurately approximation depending on the application. We may also come to the same conclusion by looking at the posterior predictions and covariance function plots in Figure~\ref{fig3_Post_part1}. From these results, some general conclusions may be drawn:

@@

\begin{figure}[!h]
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 5mm 0mm, clip]{ch5_fig5_lscale_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 10mm 0mm, clip]{ch5_fig5_magnitud_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 28mm 35mm 119mm 0mm, clip]{ch5_fig5_legend.pdf}}
\caption{Estimated lengthscale (left) and marginal variance (right) parameters of both exact GP and HSGP models, plotted versus the number of basis functions $m$ and for different values of the boundary factor $c$.}
  \label{fig5_lscale_vs_J}
\end{figure}

Additionally, there is a clear relation between the number of basis functions $m$ and the boundary factor $c$ with the lengthscale $\ell$ of the approximated function. Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} depict how these three factors interact and affect the accuracy of the HSGP approximation for a GP with square exponential covariance function and Mat\'ern ($\nu=3/2$) covariance function, respectively, and a single input dimension. More precisely, for a given GP model (with a squared exponential covariance function) with lengthscale $\ell$ and given a boundary factor $c$, Figure \ref{fig6_lscale_vs_J_vs_c} shows the minimum number of basis functions $m$ required to obtain an accurate approximation in the sense of satisfying eq. (\ref{eq_diff_covs}). Similarly,  Figure \ref{fig6_lscale_vs_J_vs_c_Matern} shows the corresponding plot for the  Mat\'ern ($\nu=3/2$) covariance function. We considered an approximation to be a sufficiently accurate when the total variation difference between the approximate and exact covariance functions, $\varepsilon$ in eq. (\ref{eq_diff_covs}), is below 1$\%$ of the total area under the curve of the exact covariance function $k$ such that
%
\begin{equation*}
 \int | k(\tau) - \tilde{k}_m(\tau)|  \,\mathrm{d}\tau  < 0.01 \int k(\tau) \,\mathrm{d}\tau,
\end{equation*}
where $\tilde{k}_m$ is the approximate covariance function with $m$ basis functions. Alternatively, these figures can be understood as providing the minimum $c$ that we should use for given $\ell$ and $m$. Of course, we may also read it as providing the minimum $\ell$ that can be approximated with high accuracy given $m$ and $c$. We obtain the following main conclusions:

@@

As stated above, Figures~\ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} provide the minimum lengthscale that can be closely approximated given $m$ and $c$. This information serves as a powerful diagnostic tool in determining if the obtained accuracy is acceptable. As the lengthscale $\ell$ controls the wigglyness of the function, it strongly influences the difficulty of obtaining accurate inference about the function from the data. Basically, if the lengthscale estimate is accurate, we can expect the HSGP approximation to be accurate as well.

Having obtained an estimate $\hat{\ell}$ of $\ell$ from the HSGP model based on prespecified $m$ and $c$, we can check whether or not $\hat{\ell}$ exceeds the minimum lengthscale provided in Figure~\ref{fig6_lscale_vs_J_vs_c} or \ref{fig6_lscale_vs_J_vs_c_Matern} (depending on which kernel is used). If $\hat{\ell}$ exceeds this recommended minimum lengthscale, the approximation is assumed to be good. If, however, $\hat{\ell}$ does not exceed recommended minimum lengthscale, the approximation may be inaccurate and $m$ should be increased or $c$ decreased. We may also use this diagnostic in an iterative procedure by starting from some initial guess of $\ell$ and initial values for $m$ and $c$, and if the estimated $\hat{\ell}$ is below the minimum lengthscale, repeat the process while increasing $m$ or decreasing $c$. As mentioned earlier, $c$ cannot be decreased too much as the lowest useful value of $c$ is restricted by the lengthscale. Thus, increasing $m$ may usually the preferred approach.

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.40, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig6_lscale_vs_J_vs_c.pdf}}
\hspace{3mm}
\subfigure{\includegraphics[scale=0.40, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig6_lscale_vs_J_vs_c_zoomin.pdf}}
\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ \hspace{0.1mm} ($c = \frac{L}{S}$) \hspace{0.1mm} and the lengthscale normalized by the half-range of the data ($\frac{\ell}{S}$), in the case of a square exponential covariance function. The right-side plot is a zoom in of the left-side plot.}
  \label{fig6_lscale_vs_J_vs_c}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.40, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig6_lscale_vs_J_vs_c_Matern.pdf}}
\hspace{3mm}
\subfigure{\includegraphics[scale=0.40, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig6_lscale_vs_J_vs_c_zoomin_Matern.pdf}}
\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ \hspace{0.1mm} ($c = \frac{L}{S}$) \hspace{0.1mm} and the lengthscale normalized by the half-range of the data ($\frac{\ell}{S}$), in the case of a Mat\`ern($\nu$=3/2) covariance function. The right-side plot is a zoom in of the left-side plot.}
  \label{fig6_lscale_vs_J_vs_c_Matern}
\end{figure}

@@

\subsection{Comparing lengthscale estimates}

In this example, we make a comparison of the lengthscale estimates
obtained from the exact GP and HSGP models. We also have a look at those recommended minimum lengthscales provided by Figure \ref{fig6_lscale_vs_J_vs_c}.
For this analysis, we will use various datasets consisting of noisy draws from a GP prior model with a squared exponential covariance function and varying lengthscale values. Different values of the number of basis functions $m$ are used when estimating the HSGP models, and the boundary factor $c$ is set to a valid and optimum value in every case.

Figure \ref{fig7_posterior_varing_lscale_part1} shows the posterior predictions of both exact GP and HSGP models fitted to those datasets. The lengthscale estimates as obtained by exact GP and HSGP models are depicted in Figure \ref{fig8_Tlscale_vs_Elscale}. As noted previously, an accurate estimate of the lengthscale can be a good indicator of a close approximation of the HSGP model to the exact GP model. Further, Figure \ref{fig9_MSE_varing_lscale} shows the root mean square error (RMSE) of the HSGP models, computed against the exact GP models, as a function of the lengthscale and number of basis functions.

Comparing the accuracy of the lengthscale in Figure~\ref{fig8_Tlscale_vs_Elscale} to the RMSE in Figure~\ref{fig9_MSE_varing_lscale}, we see that they agree closely with each other for medium lengthscales. That is, a good estimation of the lengthscale implies a small RMSE. This is no longer true for very small or large lengthscales. In small lengthscales, even very small inaccuracies  may have a strong influence on the posteriors predictions and thus on the RMSE. In large lengthscales, larger inaccuracies change the posterior predictions only little and may thus not yield large RMSEs. The dashed black line in Figure \ref{fig8_Tlscale_vs_Elscale} represents the  minimum lengthscale that can be closely approximated under the given condition, according to the results presented in Figure \ref{fig6_lscale_vs_J_vs_c}.  We observe that whenever the estimated lengthscale exceeds the minimally estimable lengthscale, the RMSE of the posterior predictions is small
(see Figure \ref{fig9_MSE_varing_lscale}). Conversely, when the estimated lengthscale is smaller than the minimally estimable one, the RMSE becomes very large.

\begin{figure}[H]
\centering
\subfigure{\includegraphics[scale=0.42, trim = 10mm 100mm 5mm 30mm, clip]{ch5_fig7_legend.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part1.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part2.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part3.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part4.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part5.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part6.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part7.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part8.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part9.pdf}}\\
\caption{Mean posterior predictions of both exact GP and HSGP models, fitted over various datasets drawn from square exponential GP models with different characteristic lengthscales ($\ell$) and same marginal variance ($\alpha$) as the data-generating functions (\textit{true function}). }
  \label{fig7_posterior_varing_lscale_part1}
\end{figure}

\begin{figure}
\begin{flushleft}
\begin{tabular}{ccc}
\multicolumn{3}{c}{ \includegraphics[scale=0.42, trim = 25mm 117mm 90mm 45mm, clip]{ch5_fig8_legend.pdf}}\\
\includegraphics[scale=0.31, trim = 0mm 24mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part1.pdf} & \hspace{-4mm}
\includegraphics[scale=0.31, trim = 25mm 24mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part2.pdf} & \hspace{-4mm} \includegraphics[scale=0.31, trim = 25mm 24mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part3.pdf} \\ 
\includegraphics[scale=0.31, trim = 0mm 2mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part4.pdf} & \hspace{-4mm} \includegraphics[scale=0.31, trim = 25mm 2mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part5.pdf} & \hspace{-4mm} \includegraphics[scale=0.31, trim = 25mm 2mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part6.pdf}
\end{tabular}
\end{flushleft}
\caption{Data-generating functional lengthscales ($\ell$), of the various datasets illustrated in Figure \ref{fig7_posterior_varing_lscale_part1}, versus the corresponding lengthscale estimates $(\hat{\ell})$ from the exact GP and HSGP models. 95\% confident intervals of the lengthscale estimates are plotted as dot lines. The different plots represent the use of different number of basis functions $m$ in the HSGP model. The dashed black line represents the recommended minimum lengthscales provided by Figure \ref{fig6_lscale_vs_J_vs_c} that can be closely approximated by the HSGP model in every case.}
  \label{fig8_Tlscale_vs_Elscale}
\end{figure}

@@

In this experiment, we analyze a synthetic dataset with $n = 250$ observations, where the true data generating process is a Gaussian process with additive noise. The data points are simulated from the model $y_i = f(x_i) + \epsilon_i$, where $f$ is a sample from a Gaussian process using the Mat{\'e}rn($\nu$=3/2) covariance function with marginal variance $\alpha=1$ and lengthscale $\ell=0.15$ at inputs values $\bm{x}=(x_1,x_2,\dots,x_n)$ with $x_i \in [-1,1]$. $\epsilon_i$ is additive Gaussian noise with standard deviation $\sigma=0.2$. We split the dataset into three parts: $155$ data points are used for fitting the model (training set), $45$ data ponts are used for the interpolation test set, and the remaining $50$ data points are used for the extrapolation test set.

>>>>

The accuracy and speed of the HSGP model depends on several interrelated factors, most notably on the number of basis functions and on the boundary condition of the Laplace eigenfunctions. Furthermore, appropriate values for these factors will depend on the degree of non-linearity of the function to be estimated, which is in turn is characterized by the length-scale of the covariance function. In this section, we analyze the effects of the number of basis functions and the boundary condition on the approximation accuracy. We present recommendations on how they should be chosen and diagnostics to check the accuracy of the obtained approximation.

Ultimately, these recommendations are based on the relationships among the number of basis functions, the boundary condition and the length-scale of the function, which depend on the particular choice of the kernel function. In this work we investigate these relationships for the square exponential and the Mat{\'e}rn ($\nu$=3/2) covariance functions in the present section, and for the periodic squared exponential covariance function in Appendix~\ref{ch5_sec_periodic}. For other kernels, the relationships will be slightly different depending on the smoothness or wigglyness of the covariance function.

@@

This inequality can be satisfied for arbritrary small $\epsilon$ provided that $L$ and $m$ are sufficiently large \cite[Theorem 1 and 4]{solin2018hilbert}. 
The specific number of basis functions $m$ needed depends on the degree of non-linearity of the function to be estimated, that is on its length-scale $\ell$, which constitutes a hyperparameter of the G. The approximation also depends on the boundary $L$ (see equations (\ref{eq_eigenvalue}), (\ref{eq_eigenfunction}), (\ref{eq_eigenvalue_multi}) and (\ref{eq_eigenfunction_multi})), which will affect its accuracy especially near the boundaries. As we will see later on, $L$ will also influence the number of basis functions required in the approximation.

@@

We start by illustrating how the number of basis functions $m$ and boundary factor $c$ influence the accuracy of the HSGP approximations individually. For this purpose, a set of noisy observations are drawn from an exact GP model with a squared exponential covariance function of length-scale $\ell=0.3$ and marginal variance $\alpha=1$, using input values from the zero-mean input domain with half-range $S=1$. Several HSGP models with varying $m$ and $c$ are fitted to this data. In this example, the length-scale and marginal variance parameters used in the HSGPs are fixed to the true values of the data-generating model. 
Figures~\ref{fig1_Post_J} and \ref{fig2_Post_L} illustrate the individual effects of $m$ and $c$, respectively, on the posterior predictive mean and standard deviation of the estimated function as well as on the covariance function itself. For $c$ fixed to a sufficiently large value, Figure \ref{fig1_Post_J} shows clearly how $m$ affects the accuracy on the approximation for both the posterior mean or uncertainty. It is seen that if the number of basis functions $m$ is too small, the estimated function tend to be overly smooth because the necessary high frequency components are missing. In general, the higher the degree of wigglyness of the function to be estimated, the larger number of  basis functions will be required. If $m$ is fixed to a sufficiently large value, Figure~\ref{fig2_Post_L} shows that $c$ affects the approximation mainly near the boundaries in mean, but along the whole domain in the standard deviation. The approximation error for the variance tends to be bigger for the variance than for the mean.

@@

Next, we analyze how the interaction effects between $m$ and $c$ affects the quality of the approximation. The length-scale and marginal variance of the covariance function will no longer be fixed but instead we compute the joint posterior distribution of the function values and the hyperparameters using the dynamic HMC algorithm implemented in Stan \citep{carpenter2017stan,betancourt2017conceptual} for both the exact GP and the HSGP models. Figure~\ref{fig3_Post_part1} shows the posterior predictive mean and standard deviation of the function as well as the covariance function obtained after fitting the model for varying $m$ and $c$. Figure \ref{fig4_MSE_vs_J} shows the root mean square error (RMSE) of the HSGP models computed against the exact GP model. Figure \ref{fig5_lscale_vs_J} shows the estimated length-scale and marginal variance for the exact GP model and the HSGP models. Looking at the RMSEs in Figure \ref{fig4_MSE_vs_J}, we can conclude that the optimal choice in terms of precision and computation time for this example would be $m = 15$ basis functions and a boundary factor between $c = 1.5$ and $c = 2.5$. Further, the less conservative choice of $m = 10$ and $c = 1.5$  could also produce a sufficiently accurately approximation depending on the application. We may also come to the same conclusion by looking at the posterior predictions and covariance function plots in Figure~\ref{fig3_Post_part1}. From these results, some general conclusions may be drawn:

@@

\begin{figure}[!h]
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 5mm 0mm, clip]{ch5_fig5_lscale_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 10mm 0mm, clip]{ch5_fig5_magnitud_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 28mm 35mm 119mm 0mm, clip]{ch5_fig5_legend.pdf}}
\caption{Estimated length-scale (left) and marginal variance (right) parameters of both exact GP and HSGP models, plotted versus the number of basis functions $m$ and for different values of the boundary factor $c$.}
  \label{fig5_lscale_vs_J}
\end{figure}

Additionally, there is a clear relation between the number of basis functions $m$ and the boundary factor $c$ with the length-scale $\ell$ of the approximated function. Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} depict how these three factors interact and affect the accuracy of the HSGP approximation for a GP with square exponential covariance function and Mat\'ern ($\nu=3/2$) covariance function, respectively, and a single input dimension. More precisely, for a given GP model (with a squared exponential covariance function) with length-scale $\ell$ and given a boundary factor $c$, Figure \ref{fig6_lscale_vs_J_vs_c} shows the minimum number of basis functions $m$ required to obtain an accurate approximation in the sense of satisfying eq. (\ref{eq_diff_covs}). Similarly,  Figure \ref{fig6_lscale_vs_J_vs_c_Matern} shows the corresponding plot for the  Mat\'ern ($\nu=3/2$) covariance function. We considered an approximation to be a sufficiently accurate when the total variation difference between the approximate and exact covariance functions, $\varepsilon$ in eq. (\ref{eq_diff_covs}), is below 1$\%$ of the total area under the curve of the exact covariance function $k$ such that
%
\begin{equation*}
 \int | k(\tau) - \tilde{k}_m(\tau)|  \,\mathrm{d}\tau  < 0.01 \int k(\tau) \,\mathrm{d}\tau,
\end{equation*}
where $\tilde{k}_m$ is the approximate covariance function with $m$ basis functions. Alternatively, these figures can be understood as providing the minimum $c$ that we should use for given $\ell$ and $m$. Of course, we may also read it as providing the minimum $\ell$ that can be approximated with high accuracy given $m$ and $c$. We obtain the following main conclusions:

@@

As stated above, Figures~\ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} provide the minimum length-scale that can be closely approximated given $m$ and $c$. This information serves as a powerful diagnostic tool in determining if the obtained accuracy is acceptable. As the length-scale $\ell$ controls the wigglyness of the function, it strongly influences the difficulty of obtaining accurate inference about the function from the data. Basically, if the length-scale estimate is accurate, we can expect the HSGP approximation to be accurate as well.

Having obtained an estimate $\hat{\ell}$ of $\ell$ from the HSGP model based on prespecified $m$ and $c$, we can check whether or not $\hat{\ell}$ exceeds the minimum length-scale provided in Figure~\ref{fig6_lscale_vs_J_vs_c} or \ref{fig6_lscale_vs_J_vs_c_Matern} (depending on which kernel is used). If $\hat{\ell}$ exceeds this recommended minimum length-scale, the approximation is assumed to be good. If, however, $\hat{\ell}$ does not exceed recommended minimum length-scale, the approximation may be inaccurate and $m$ should be increased or $c$ decreased. We may also use this diagnostic in an iterative procedure by starting from some initial guess of $\ell$ and initial values for $m$ and $c$, and if the estimated $\hat{\ell}$ is below the minimum length-scale, repeat the process while increasing $m$ or decreasing $c$. As mentioned earlier, $c$ cannot be decreased too much as the lowest useful value of $c$ is restricted by the length-scale. Thus, increasing $m$ may usually the preferred approach.

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.40, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig6_lscale_vs_J_vs_c.pdf}}
\hspace{3mm}
\subfigure{\includegraphics[scale=0.40, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig6_lscale_vs_J_vs_c_zoomin.pdf}}
\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ \hspace{0.1mm} ($c = \frac{L}{S}$) \hspace{0.1mm} and the length-scale normalized by the half-range of the data ($\frac{\ell}{S}$), in the case of a square exponential covariance function. The right-side plot is a zoom in of the left-side plot.}
  \label{fig6_lscale_vs_J_vs_c}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.40, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig6_lscale_vs_J_vs_c_Matern.pdf}}
\hspace{3mm}
\subfigure{\includegraphics[scale=0.40, trim = 0mm 0mm 5mm 0mm, clip]{ch5_fig6_lscale_vs_J_vs_c_zoomin_Matern.pdf}}
\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ \hspace{0.1mm} ($c = \frac{L}{S}$) \hspace{0.1mm} and the length-scale normalized by the half-range of the data ($\frac{\ell}{S}$), in the case of a Mat\`ern($\nu$=3/2) covariance function. The right-side plot is a zoom in of the left-side plot.}
  \label{fig6_lscale_vs_J_vs_c_Matern}
\end{figure}

@@

\subsection{Comparing length-scale estimates}

In this example, we make a comparison of the length-scale estimates
obtained from the exact GP and HSGP models. We also have a look at those recommended minimum length-scales provided by Figure \ref{fig6_lscale_vs_J_vs_c}.
For this analysis, we will use various datasets consisting of noisy draws from a GP prior model with a squared exponential covariance function and varying length-scale values. Different values of the number of basis functions $m$ are used when estimating the HSGP models, and the boundary factor $c$ is set to a valid and optimum value in every case.

Figure \ref{fig7_posterior_varing_lscale_part1} shows the posterior predictions of both exact GP and HSGP models fitted to those datasets. The length-scale estimates as obtained by exact GP and HSGP models are depicted in Figure \ref{fig8_Tlscale_vs_Elscale}. As noted previously, an accurate estimate of the length-scale can be a good indicator of a close approximation of the HSGP model to the exact GP model. Further, Figure \ref{fig9_MSE_varing_lscale} shows the root mean square error (RMSE) of the HSGP models, computed against the exact GP models, as a function of the length-scale and number of basis functions.

Comparing the accuracy of the length-scale in Figure~\ref{fig8_Tlscale_vs_Elscale} to the RMSE in Figure~\ref{fig9_MSE_varing_lscale}, we see that they agree closely with each other for medium length-scales. That is, a good estimation of the length-scale implies a small RMSE. This is no longer true for very small or large length-scales. In small length-scales, even very small inaccuracies  may have a strong influence on the posteriors predictions and thus on the RMSE. In large length-scales, larger inaccuracies change the posterior predictions only little and may thus not yield large RMSEs. The dashed black line in Figure \ref{fig8_Tlscale_vs_Elscale} represents the  minimum length-scale that can be closely approximated under the given condition, according to the results presented in Figure \ref{fig6_lscale_vs_J_vs_c}.  We observe that whenever the estimated length-scale exceeds the minimally estimable length-scale, the RMSE of the posterior predictions is small
(see Figure \ref{fig9_MSE_varing_lscale}). Conversely, when the estimated length-scale is smaller than the minimally estimable one, the RMSE becomes very large.

\begin{figure}[H]
\centering
\subfigure{\includegraphics[scale=0.42, trim = 10mm 100mm 5mm 30mm, clip]{ch5_fig7_legend.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part1.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part2.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part3.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part4.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part5.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part6.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part7.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part8.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part9.pdf}}\\
\caption{Mean posterior predictions of both exact GP and HSGP models, fitted over various datasets drawn from square exponential GP models with different characteristic length-scales ($\ell$) and same marginal variance ($\alpha$) as the data-generating functions (\textit{true function}). }
  \label{fig7_posterior_varing_lscale_part1}
\end{figure}

\begin{figure}
\begin{flushleft}
\begin{tabular}{ccc}
\multicolumn{3}{c}{ \includegraphics[scale=0.42, trim = 25mm 117mm 90mm 45mm, clip]{ch5_fig8_legend.pdf}}\\
\includegraphics[scale=0.31, trim = 0mm 24mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part1.pdf} & \hspace{-4mm}
\includegraphics[scale=0.31, trim = 25mm 24mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part2.pdf} & \hspace{-4mm} \includegraphics[scale=0.31, trim = 25mm 24mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part3.pdf} \\ 
\includegraphics[scale=0.31, trim = 0mm 2mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part4.pdf} & \hspace{-4mm} \includegraphics[scale=0.31, trim = 25mm 2mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part5.pdf} & \hspace{-4mm} \includegraphics[scale=0.31, trim = 25mm 2mm 7mm 10mm, clip]{ch5_fig8_Tlscale_vs_Elscale_part6.pdf}
\end{tabular}
\end{flushleft}
\caption{Data-generating functional length-scales ($\ell$), of the various datasets illustrated in Figure \ref{fig7_posterior_varing_lscale_part1}, versus the corresponding length-scale estimates $(\hat{\ell})$ from the exact GP and HSGP models. 95\% confident intervals of the length-scale estimates are plotted as dot lines. The different plots represent the use of different number of basis functions $m$ in the HSGP model. The dashed black line represents the recommended minimum length-scales provided by Figure \ref{fig6_lscale_vs_J_vs_c} that can be closely approximated by the HSGP model in every case.}
  \label{fig8_Tlscale_vs_Elscale}
\end{figure}

@@

In this experiment, we analyze a synthetic dataset with $n = 250$ observations, where the true data generating process is a Gaussian process with additive noise. The data points are simulated from the model $y_i = f(x_i) + \epsilon_i$, where $f$ is a sample from a Gaussian process using the Mat{\'e}rn($\nu$=3/2) covariance function with marginal variance $\alpha=1$ and length-scale $\ell=0.15$ at inputs values $\bm{x}=(x_1,x_2,\dots,x_n)$ with $x_i \in [-1,1]$. $\epsilon_i$ is additive Gaussian noise with standard deviation $\sigma=0.2$. We split the dataset into three parts: $155$ data points are used for fitting the model (training set), $45$ data ponts are used for the interpolation test set, and the remaining $50$ data points are used for the extrapolation test set.

The exact GP model for fitting and predicting this simulated dataset $\bm{y}$ can be written as follows,
%
\begin{align*}\label{ch5_eq_latentgp_simudata1}
\begin{split}
\bm{y} &= \bm{f} + \bm{\epsilon} \\
\bm{\epsilon} &\sim \Normal(\bm{0}, \sigma^2 \bm{I}) \\
f(x) &\sim \GP(0, k(x, x', \theta)),
\end{split}
\end{align*}
%
where $\bm{f}=\{f(x_i)\}_{i=1}^n$ represents the underlying function at the input values $x_i$, and $\bm{\epsilon}$ is the Gaussian noise term with variance $\sigma^2$, with $\bm{I}$ representing the identity matrix. The function $f:{\rm I\!R} \to {\rm I\!R}$ is a GP prior with the Mat{\'e}rn($\nu$=3/2) covariance function $k$, which depends on the inputs $\bm{x}$ and hyperparameters $\theta=\{\alpha,\ell\}$. The hyperparameters $\alpha$ and $\ell$ represent the marginal variance and length-scale, respectively, of the GP process. Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ is multivariate Gaussian distributed with covariance matrix $\bm{K}$, where $K_{ij}=k(x_i,x_j,\theta)$, with $i,j=1,2,\dots,n$.
